Quasi-universal acceptance of basic science in the US

### 2024-09-10 paper.tex
before final edits Hugo, pre-submission

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Universal acceptance of (nearly all of) basic science in the US},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Universal acceptance of (nearly all of) basic science in the US}
\author{\textsuperscript{}}
\date{}

\shorttitle{Acceptance of basic science}

@@

How much do people trust and accept science? The recent COVID pandemic has highlighted the importance of this question, with trust in science being the best predictor of people's proclivity to follow health guidelines (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021).

Unfortunately, trust in science is far from being at ceiling (Wellcome Global Monitor, 2018, 2020), and there are suggestions that it has recently been dropping in some countries (Algan et al., 2021), including the US (Brian \& Tyson, 2023), where it is also increasingly politicized (Gauchat, 2012; Krause, Brossard, Scheufele, Xenos, \& Franke, 2019; Li \& Qian, 2022).

@@

@@

In a series of four pre-registered studies (n = 782), we asked participants about basic (studies 1, 2 and 3) and recent/advanced (study 4) science knowledge questions, informed them of the scientific consensus, and aked if they accept it. Of particular interest were people who are low in TiS, and CTists. Our main hypotheses were:

@@

In study 1, we plainly inform participants of the scientific consensus and ask if they accept it. We use a subset of questions on basic science knowledge that have been used in numerous public opinion surveys (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). In study 2, we remove a problematic science question and present people with a more elaborated explanation of the scientific consensus, as well as additional sources. This allows us to remove the issue that participants might simply not have trusted us to correctly report the scientific consensus in study 1. Study 3 follows essentially the same design as study 2, but is run on a vaccine-skeptic sample. Further, we provided participants with an explicit option to revise their answer, in case they rejected the consensus. We also asked follow-up questions on why participants accepted the scientific consensus, namely whether they thought it was because they trust scientists or because they verified independently. Study 4 uses the same design as study 3, but using a set of more recent, advance science questions that participants are less likely to have encountered before, and that they would be less likely to be able to verify (or even understand) themselves. Just as study 3, study 4 is run on a sample of vaccine-skeptic participants.

@@

@@

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions (in study 1 they were randomly selected from a pool of 11 questions) in random order. After each question, participants were presented with an answer reflecting the scientific consensus. In study 2 and 3, participants additionally saw a short explanation which we wrote, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In study 4, we provided only two links and no explanation. Participants were asked to choose whether they accept the answer or not, before proceeding to the next question. Participants then answered questions on conspiracy thinking, conspiracy beliefs and trust in science.

In study 2, 3 and 4, we presented participants with open-ended questions to explain cases where they rejected the scientific consensus. In study 3 and 4, we additionally gave participants the option to change their answer and accept the scientific consensus. Finally, at the end of study 3 and 4, we asked participants: ``For the questions in which you agreed with the scientific consensus, would you say that\ldots?'' The answer options were: (i) ``You mostly agree with the consensus because, on that question, you trust scientists'', (ii) ``You mostly agree with the consensus because you have been able to independently verify it'', and (iii) ``Other'', with a text box for participants to explain. Participants who selected ``You mostly agree with the consensus because you have been able to independently verify it'', were asked the open-ended follow-up question: ``Could you please tell us how you independently verified the information?''.

@@

We recruited a total of 782 (194 in study 1, 190 in study 2, 200 in study 3, 198 in study 4) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for study 1 and 2 were convenience samples, study 3 and 4 were run on a sample holding vaccine-skeptic beliefs (see OSM for details).

\subsection{Materials}\label{materials}

Table \ref{tab:knowledge} shows all questions and their answer options used in the four studies.

@@

Our main measure of conspiracy belief was asking participants about specific conspiracy theories. We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) (Pennycook, Binnendyk, \& Rand, 2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

As a robustness check, we also used established scales on conspiracy thinking, namely the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the OSM.

@@

\begin{longtable}[t]{>{}r>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:conspiracy}Conspiracy items}\\
\toprule
1 & The Apollo moon landings never happened and were staged in a Hollywood film studio.\\
2 & A cure for cancer was discovered years ago, but this has been suppressed by the pharmaceutical industry and the U.S. Food and Drug Administration (FDA).\\
3 & The spread of certain viruses and/or diseases is the result of the deliberate, concealed efforts of vested interests.\\
4 & The claim that the climate is changing due to emissions from fossil fuels is a hoax perpetrated by corrupt scientists who want to spend more taxpayer money on climate research.\\
5 & The Earth is flat (not spherical) and this fact has been covered up by scientists and vested interests.\\
\addlinespace
6 & There is a causal link between vaccination and autism that has been covered up by the pharmaceutical industry.\\
7 & In the 1950s and 1960s more than 100 million Americans received a polio vaccine contaminated with a potentially cancer-causing virus.\\
8 & Proof of alien contact is being concealed from the public.\\
9 & Hydroxychloroquine has been demonstrated to be a safe and effective treatment of COVID and this information is being suppressed.\\
10 & Dinosaurs never existed, evolution is not real, and scientists have been faking the fossil record.\\
\bottomrule
\end{longtable}
\endgroup{}

Our main item for measuring trust in science was selected from the Wellcome Global Monitor survey: ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''. We also included two additional trust questions, one also from the Wellcome Global Monitor (WGM) survey (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), the other from the Pew research center (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''). We selected these items so that we could compare the ratings in our sample to global survey results. The WGM survey has been administered in over 140 countries and included over 140000 respondents. The Pew question has recently been used by a world-wide many labs study in 67 countries with 71417 respondents (Cologna et al., 2024).

@@

\begin{figure}
\centering
\includegraphics{output/figures/summary-plot.pdf}
\caption{\label{fig:summary-plot}Points represent the average share of acceptance as a function of the \textbf{A} level of trust in science (``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''); \textbf{B} average conspiracy thinking (CMQ, 5 items on a scale from 0 to 100); \textbf{C} belief in a specific conspiracy theory. Participants were asked to rate their belief in the conspiracy on a scale from 1 to 9, with the labels: ``1 - completly false, 5 - unsure, 9 - completely true''. Averages are only based on participants who believed at least to some degree in the conspiracy theory. We consider a believer as anyone scoring higher than the scale midpoint, i.e.~\textgreater5. Number labels in plots represent the number of participants for the corresponding data point.}
\end{figure}

Across all studies we find that almost all participants trust basic and recent/advance science facts almost all the time. That is true even for participants who expressed low general trust in science, and those who have a tendency to conspiracy think or hold science-related conspiracy beliefs (Fig. \ref{fig:summary-plot}).

Our results suggest that the more people tend to trust science, the more they tend to know about science and accept the scientific consensus. In study 2 and 3, we find a positive but small correlation between trust in science and both science knowledge (study 1: r = 0.29, p \textless{} .001; study 2: r = 0.28, p \textless{} .001), and acceptance of scientific consensus (study 1: r = 0.27, p \textless{} .001; study 2: r = 0.30, p \textless{} .001). In study 3 and 4--both run on vaccine-skeptic samples--we find no statistically significant correlation between science knowledge and trust in science (study 3: r = 0.14, p = 0.100; study 4: r = 0.10, p = 0.144), but a small positive correlation between acceptance of scientific consensus and trust in science (study 3: r = 0.20, p = 0.019; study 4: r = 0.16, p = 0.029).

As for conspiracy beliefs, results are mixed. In studies 1 and 2, we find a small negative correlation between conspiracy thinking and both science knowledge (study 1: r = -0.38, p \textless{} .001; study 2: r = -0.40, p \textless{} .001), and acceptance of scientific consensus (study 1: r = -0.33, p \textless{} .001; study 2: r = -0.37, p \textless{} .001). However, in study 3 and 4, we do not find a statistically significant correlation with neither science knowledge (study 3: r = -0.16, p = 0.055; study 4: r = -0.02, p = 0.742), and between nor acceptance of scientific consensus (study 3: r = -0.02, p = 0.788; study 4: r = -0.05, p = 0.498).

Overall, the correlations we find are relatively weak, which might be partly due to ceiling effects. This was the case in particular for acceptance: In all studies, in the vast majority of cases, participants accepted accepted science facts. The absence of correlations between conspiracy beliefs and science knowledge or acceptance in studies 3 and 4 might be explained by our sample selection. For these studies, we recruited samples scoring considerably higher in conspiracy belief than the convenience samples of study 1 and 2 (see OSM). The presence of effects in study 1 and 2, and the absence in study 3 and 4 might be because differences between moderate and high conspiracy thinking might not matter, while, differences between low and moderate/high conspiracy thinking do matter.

Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. Only in study 4 do we find evidence\footnote{For this analysis, we restricted our data to those cases in which participants initially gave the wrong answer.} that changing one's mind towards the scientific consensus is associated with (more) trust in science (study 1: r = 0.06, p = 0.387; study 2: r = 0.16, p = 0.051; study 3: r = 0.04, p = 0.619; study 4: r = 0.15, p = 0.037) and only in study 2 evidence that it is associated with (less) conspiracy thinking (study 1: r = -0.14, p = 0.061; study 2: r = -0.22, p = 0.006; study 3: r = -0.04, p = 0.631; study 4: r = -0.05, p = 0.455). In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: 1.6 \%; study 2: 0.5 \%; study 3: 0.7 \%; study 4: 5.7 \%).

@@

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
NA & 255.00 & 70.1\% & 95.00\\
Personal convictions & 37.00 & 10.2\% & 27.00\\
Mistake & 30.00 & 8.2\% & 26.00\\
No justification & 25.00 & 6.9\% & 18.00\\
Not convinced & 12.00 & 3.3\% & 8.00\\
Religious Beliefs & 5.00 & 1.4\% & 5.00\\
\bottomrule
\end{tabular}

Why did participants accept the scientific consensus? In studies 3 and 4, we had asked participants about cases where they agreed with the scientific consensus. A total of 320 (study 3: 122; study 4: 198) participants answered this question. Of these, 41.8\% in study 3 and 36.4\% in study 4 said they accepted the scientific consensus because they trust scientists, while 47.5\% in study 3 and 47\% in study 4 said they independently verified the fact. We asked all participants who gave this answer to explain how they did it (answers can be found in the OSM). 10.7\% in study 3 and 16.7\% in study 4 answered with other ``other'' and gave an open-ended explanation (see OSM).

@@

Scholars have long interpreted the lack of basic science knowledge as the root cause of distrust in and more generally unfavorable attitudes towards science, an explanation known under the term ``deficit-model'' (Durant et al., 1989; Miller, 1998, 2004). In line with proponents of the deficit-model, we find that people indeed err even on very basic science questions. However, we also find that almost everyone instantly accepted the scientific consensus once informed about it. Moreover, the same was true for non-basic science questions. This calls into question the deficit model: If participants trust science on basic (and non-basic) questions---as suggests immediate acceptance of the scientific consensus---then unfavorable attitudes towards science must have other explanations than a mere lack of knowledge. This argument is supported by past research. In a seminal meta-analysis, @allumScienceKnowledgeAttitudes2008 found correlations between science knowledge and general attitudes towards science to be at best small and largely context dependent {[}@allumScienceKnowledgeAttitudes2008{]}. A more recent report by the National Academies of Sciences, Engineering, and Medicine states that science knowledge is ``unlikely to substantially affect attitudes on scientific issues'' (National Academies of Sciences, 2017).

>>>>

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quasi-universal acceptance of basic science in the US},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Quasi-universal acceptance of basic science in the US}
\author{\textsuperscript{}}
\date{}

\shorttitle{Does anyone mistrust basic science?}

@@

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (ref) or vaccination (ref) to following recommendations during COVID (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021), which suggests that trust in science was the most important predictor of these behaviors).

Unfortunately, people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia (Wellcome Global Monitor, 2018, 2020). Moreover, trust in science has recently been declining in some countries (Algan et al., 2021; although see Wellcome Global Monitor, 2021), including the US (Brian \& Tyson, 2023), where it is also increasingly polarizing (Gauchat, 2012; Krause, Brossard, Scheufele, Xenos, \& Franke, 2019; Li \& Qian, 2022). Rejection of science can take the more extreme form of conspiracy theories, many of which question the scientific consensus on vaccination, climate change, and even the shape of the Earth.

What does this apparent rejection of science actually entail? Do people who say they do not trust science much, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it's possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus (e.g.~on vaccination, climate change, etc., for review, see Van Stekelenburg, Schaap, Veling, Van 'T Riet, \& Buijzen, 2022; see also Većkalov et al., 2024). These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

@@

{[}introduce SK as: that gives us a lower bound of how much people trust basic science{]}

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted correct answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants' trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (OSF)). The differences between the four studies are summarized presently, and the methods are detailed below.

\subsubsection{Materials}\label{materials}

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g.~``Are electrons smaller, larger, or the same size as atoms? {[}Smaller; Same size; Larger{]}''), supplemented by a `trick' question (``Where do trees mainly draw the materials with which they create their mass? {[}Earth; Water; Air{]}''; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g.~``How many more glial cells are there in the brain in comparison with neurons, as established in 2016? {[}The same amount; Twice as many; Tenth as many{]}''; correct answer: The same amount).

\subsubsection{Presentation of the scientific consensus}\label{presentation-of-the-scientific-consensus}

In Study 1, we simply told participants that they would be provided with the scientifically consensual answer. However, for participants to accept this answer, they must not only trust science, but also trust us (that we are presenting them with the actual scientifically consensual answer). To remove this issue, in Studies 2 to 3, we presented participants with a short explanation of the correct answer, as well as links to three sources per answer (e.g.~Wikipedia, National Geographic or NASA). In Study 4, as the topics were more complex, we did not provide an explanation, but we provided two sources per answer.

\subsubsection{Measure of acceptance of the scientific consensus}\label{measure-of-acceptance-of-the-scientific-consensus}

In Study 1, we simply look at whether participants say that they accept the scientifically consensual answer. In the subsequent studies, we asked participants to explain why they disagreed with the scientific consensus. This revealed that a number of participants had made a mistake (misunderstanding, selecting the wrong answer). As a result, in Studies 3 and 4, participants who have indicated that they rejected the scientifically consensual answer were offered the option to revise their answer, or to keep rejecting it.

\subsubsection{Additional questions}\label{additional-questions}

In Studies 3 and 4, we attempted to understand how some people who say they do not trust science still accept scientifically consensual answers, by asking them whether they accepted the answers on the basis of trust in science or because they had independently verified them.

\subsubsection{Samples}\label{samples}

Studies 1 and 2 were conducted on the standard sample of US participants recruited on the platform Prolific Academic. In order to increase the share of participants with low trust in science, and who endorse conspiracy theories, Studies 3 and 4 used the same platform, but only recruited participants who had declared previously being skeptical of vaccination.

\subsubsection{Hypotheses}\label{hypotheses}

The main goal of the present studies is descriptive: to find out whether participants who report not trusting science, or who believe in conspiracy theories, still accept most well-established scientific facts. However, we also tested two directional hypotheses (pre-registered as research questions in the Study 1):

@@

@@

\subsection{Deviations from preregistration}\label{deviations-from-preregistration}

For study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. The rationale was to see if trust in science or conspiracy belief were associated with a change of mind towards the consensus. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM\footnote{Only in study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (study 1: r = 0.06, p = 0.387; study 2: r = 0.16, p = 0.051; study 3: r = 0.04, p = 0.619; study 4: r = 0.15, p = 0.037) and only in study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (study 1: r = -0.14, p = 0.061; study 2: r = -0.22, p = 0.006; study 3: r = -0.04, p = 0.631; study 4: r = -0.05, p = 0.455).}. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting simple, unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

@@

After providing their consent to participate in the study, participants were given an attention check ``While watching the television, have you ever had a fatal heart attack?'' {[}1-6; 1 = Never, 6 = Often{]}. All participants who did not answer ``1 = Never'' were excluded. Participants then read the following instructions:``We will ask you 10 questions about science. After each question, we will provide you with the scientifically consensual answer and ask whether you accept it.'' Next, participants answered a set of 10 basic science questions (in Study 1 they were randomly selected from a pool of 11 questions, as one `trick' question had been added) in random order. After each question, participants were presented with an answer reflecting the scientific consensus, and asked whether they accepted it. In Studies 2 and 3, participants additionally saw a short explanation, partly based on explanations generated by ChatGPT, and three links to authoritative sources supporting the answer. In Study 4, we provided only two links and no explanation. Participants then answered questions on conspiracy thinking, conspiracy beliefs, and trust in science.

In Studies 2, 3, and 4, we presented participants with open-ended questions so they could explain their rejection of the scientific consensus. In Studies 3 and 4, we additionally gave participants the option to change their answer and accept the scientific consensus. Finally, at the end of Studies 3 and 4, we asked participants: ``For the questions in which you agreed with the scientific consensus, would you say that\ldots?'' The answer options were: (i) ``You mostly agree with the consensus because, on that question, you trust scientists'', (ii) ``You mostly agree with the consensus because you have been able to independently verify it'', and (iii) ``Other'', with a text box for participants to explain. Participants who selected ``You mostly agree with the consensus because you have been able to independently verify it'', were asked the open-ended follow-up question: ``Could you please tell us how you independently verified the information?''.

@@

After removing failed attention checks, we were left with a total sample size of 782 (194 in Study 1, 6 failed attention checks; 190 in Study 2, 11 failed attention checks; 200 in Study 3, no failed attention checks; 198 in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows to select participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ``Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: {[}For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say''{]}. We selected particpants who answered''Against''.
\item
  ``Have you received a coronavirus (COVID-19) vaccination? {[}Yes (at least one dose); No; Prefer not to answer{]}''. We select only people who answered ``No''.
\item
  ``On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. {[}1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say{]}''. We select only people who answered `1', `2', or `3'.
\end{enumerate}

\subsubsection{Scientific facts}\label{scientific-facts}

Studies 1 to 3 used 10 facts drawn from widely used questionnaires about science knowledge (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). A `trick' question was added in Study 1 and removed as its wording proved unclear. Study 4 used 10 more recent scientific discoveries. Table \ref{tab:knowledge} shows all questions and their answer options used in the four studies.

@@

\subsubsection{Conspiracy beliefs}\label{conspiracy-beliefs}

We selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) (Pennycook, Binnendyk, \& Rand, 2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

\subsubsection{Conspiracy thinking}\label{conspiracy-thinking}

We used the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the OSM.

@@

\begin{ThreePartTable}
\begin{TableNotes}[para]
\item \textit{Note: } 
\item Participants were asked to rate their belief in the conspiracy on a scale from 1 to 9, with the labels: `1 - completely false, 5 - unsure, 9 - completely true`.
\end{TableNotes}
\begin{longtable}[t]{>{\raggedleft\arraybackslash}p{3em}>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:conspiracy}Conspiracy items}\\
\toprule
1 & The Apollo moon landings never happened and were staged in a Hollywood film studio.\\
2 & A cure for cancer was discovered years ago, but this has been suppressed by the pharmaceutical industry and the U.S. Food and Drug Administration (FDA).\\
3 & The spread of certain viruses and/or diseases is the result of the deliberate, concealed efforts of vested interests.\\
4 & The claim that the climate is changing due to emissions from fossil fuels is a hoax perpetrated by corrupt scientists who want to spend more taxpayer money on climate research.\\
5 & The Earth is flat (not spherical) and this fact has been covered up by scientists and vested interests.\\
\addlinespace
6 & There is a causal link between vaccination and autism that has been covered up by the pharmaceutical industry.\\
7 & In the 1950s and 1960s more than 100 million Americans received a polio vaccine contaminated with a potentially cancer-causing virus.\\
8 & Proof of alien contact is being concealed from the public.\\
9 & Hydroxychloroquine has been demonstrated to be a safe and effective treatment of COVID and this information is being suppressed.\\
10 & Dinosaurs never existed, evolution is not real, and scientists have been faking the fossil record.\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys (Wellcome Global Monitor, 2018, 2020): ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''. We chose this question as it seemed to be the most general one. In the appendix, we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), and one from the Pew Research Center (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''; see e.g. Funk, Johnson, and Hefferon (2019) and used by e.g. Cologna et al. (2024)). We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are generally highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures (with two exceptions: In Study 3, we find no correlation between acceptance and the Pew question; In Study 4, we find a correlation between knowledge and both alternative trust measures, but not with our main measure; see ESM).

@@

\begin{figure}
\centering
\includegraphics{output/figures/summary-plot.pdf}
\caption{\label{fig:summary-plot}Points represent the average share of acceptance and numbers the absolute count of participants as a function of: \textbf{A} the level of trust in science (``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''); \textbf{B} the average conspiracy thinking (CMQ, five items on a scale from 0 to 100); \textbf{C} the belief in specific conspiracy theories (i.e.~participants who answered 9, ``completely true'', for a given theory, see Table \ref{tab:conspiracy} for the list of the theories).}
\end{figure}

The main outcome of interest is acceptance of the scientifically consensual facts presented. Overall, acceptance was very high (aggregating across all studies: 95.1 \%; Study 1: 93 \%; Study 2: 98 \%; Study 3: 98 \%; Study 4: 91 \%). Note that this includes both participants who had previously correctly answered the knowledge question, and participants who changed their mind when presented with the scientific consensus. In Studies 3 \& 4, we gave participants a second chance in case they had initially rejected the consensus, which slightly increased acceptance rates in those studies (initial acceptance in Study 3: 96 \%; in Study 4: 86 \%).

As shown in Figure \ref{fig:summary-plot}, these very high rates of acceptance hold for: participants who they do not trust science at all (5.7 \% of participants, acceptance rate of 87.3 \%), participants who rank in the top two deciles of the conspiracy thinking scale (28.8 \% of participants, average acceptance rate of 93.2 \%), participants who consider as ``completely true'' (the maximum of the 9-point scale) conspiracy theories stating that the earth is flat (5.8 \% of participants, acceptance rate of 87.2 \%), or that climate change due to fossil emissions is a hoax (12 \% of participants, acceptance rate of 91.7 \%). This doesn't merely reflect science knowledge: aggregated across all studies--participants only correctly answered 65.8 \% of the questions (Study 1: 74 \%; Study 2: 79 \%; Study 3: 75 \%; Study 4: 36 \%) before they were provided with the scientifically consensual answer. Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: 1.6 \%; study 2: 0.5 \%; study 3: 0.7 \%; study 4: 5.7 \%).

\begin{verbatim}
## # A tibble: 1 x 2
##   mean_acceptance mean_knowledge
##             <dbl>          <dbl>
## 1           0.805          0.416
\end{verbatim}

Regarding H1, we find a consistent association between trust in science and acceptance of the scientific consensus (Studies 1: r = 0.27, p \textless{} .001; 2: r = 0.30, p \textless{} .001; 3: r = 0.20, p = 0.019; 4: r = 0.16, p = 0.029), but a less consistent relation between trust in science and science knowledge (Studies 1: r = 0.29, p \textless{} .001; 2: r = 0.28, p \textless{} .001; 3: r = 0.14, p = 0.100; study 4: r = 0.10, p = 0.144).

Regarding H2, the results are mixed both for the relation between conspiracy beliefs (measured as the average acceptance of all the conspiracy beliefs) and acceptance of the scientific consensus (Studies 1: r = -0.33, p \textless{} .001; 2: r = -0.37, p \textless{} .001; 3: r = -0.02, p = 0.788; 4: r = -0.05, p = 0.498) as well as science knowledge (Studies 1: r = -0.38, p \textless{} .001; study 2: r = -0.40, p \textless{} .001; 3: r = -0.16, p = 0.055; 4: r = -0.02, p = 0.742).

@@

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
Not convinced & 158.00 & 43.4\% & 70.00\\
No justification & 103.00 & 28.3\% & 51.00\\
Personal convictions & 43.00 & 11.8\% & 31.00\\
Mistake & 42.00 & 11.5\% & 34.00\\
Religious Beliefs & 18.00 & 4.9\% & 13.00\\
\bottomrule
\end{tabular}

Why did participants accept the scientific consensus? In Studies 3 and 4--the vaccine skeptic samples--we had asked participants about cases where they agreed with the scientific consensus. A total of 320 (Study 3: 122; Study 4: 198) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Study 3: \texttt{exp3\_descriptives\$acceptance\_by\_reason{[}{[}"independent\ verification"{]}{]}\$share}; Study 4: 47\% ), than there were participants saying it was because they trust scientists (Study 3: 41.8\%; Study 4: 36.4\%). We asked participants who said they verified independently to explain how they did it (answers can be found in the ESM). 10.7\% in study 3 and 16.7\% in study 4 answered with other ``other'' and gave an open-ended explanation (see ESM).

In an exploratory analysis, we ran linear regressions to test whether there are differences between these two groups. Perhaps unsurprisingly, those who said they accepted consensus because of trust in scientists reported to trust science more (Study 3: mean = NA; \(\hat{\beta}_{\text{Trust}}\) = 0.19, = 0.330 on a scale from 1 to 4; Study 4: mean = 2.92; \(\hat{\beta}_{\text{Trust}}\) = 0.46, \textless{} .001) than those who said they verified independently (Study 3 mean = NA; Study 4 mean = 2.45). We did not find a difference regarding acceptance (Study 3: \(\hat{\beta}_{\text{Acceptance}}\) = 0.01, = 0.523 on a scale from 0 to 1; Study 4: \(\hat{\beta}_{\text{Acceptance}}\) = 0.04, = 0.116). In study 4, where we used facts that participants were unlikely to have encountered before, we tracked whether people clicked on the source links we provided--a behavior that you would expect from people who report verifying facts independently. On average, participants clicked only clicked on 1.36 links (out of 20 possible clicks) and there was no difference between the two groups (\(\hat{\beta}_{\text{Clicks}}\) = -1.09, = 0.062).

@@

Scholars have long interpreted the lack of basic science knowledge as the root cause of distrust in and more generally unfavorable attitudes towards science, an explanation known under the term ``deficit-model'' (Durant et al., 1989; Miller, 1998, 2004). In line with proponents of the deficit-model, we find that people indeed err even on very basic science questions. However, we also find that almost everyone accepted the scientific consensus once informed about it. Moreover, the same was true for non-basic science questions. This calls into question the deficit model: If participants trust science on basic (and non-basic) questions---as suggests immediate acceptance of the scientific consensus---then unfavorable attitudes towards science must have other explanations than a mere lack of knowledge. This argument is supported by past research. In a seminal meta-analysis, @allumScienceKnowledgeAttitudes2008 found correlations between science knowledge and general attitudes towards science to be at best small and largely context dependent {[}@allumScienceKnowledgeAttitudes2008{]}. A more recent report by the National Academies of Sciences, Engineering, and Medicine states that science knowledge is ``unlikely to substantially affect attitudes on scientific issues'' (National Academies of Sciences, 2017).

How to make sense of the detachment of basic science knowledge from attitudes towards science? One explanation might be that most scientific findings are very abstract. For example, to most people, it will not be relevant for their daily lives to know that the speed of light is 299'792'458 meters per second. It is therefore easy to simply accept it, all the more if it appears to be consensual within the population of scientists. In other words, people might hold on to most scientific findings as reflective beliefs---beliefs that have little effect on our behaviors (Mercier, 2021). It is probably when it becomes tangible that science shape's people's attitudes. This can happen, for example, when science appears to support conclusions in favor of certain policies or moral positions, as for example in the case of anthropogenic climate change or vaccines.
